---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# homelessness


The goal of homelessness is to provide useful datasets on homelessness and other characteristics for researchers. The package includes 7 data sets in total: gdp, gini, homeless, population, rentincome, unemployment, and aggregated (which combines the first 6 data sets into 1 convenient data frame).

Please see the "markdown file and raw data" folder for access to the code used to collect all of the data. The R Markdown file "Final Project.Rmd" includes the code as well as links to the various APIs and organizations from which data was drawn.

## Example

This is a basic example which shows you how to calculate the total homeless population in the United States for each year.

```{r example}
library(homelessness)
library(tidyverse)
data("aggregated")
aggregated %>% group_by(year) %>% summarize(total_homeless_pop = sum(number_homeless, na.rm = TRUE))
```

To get more information about each data set, call the help function.
```{r message=FALSE}
help(gini)
```

